---
title: "BIOS735 Final Project Report"
author: "Group 5: Anil Anderson, Arthi Hariharan, David Hu, Álvaro Quijano, Yumei Yang"
date: "5/1/2024"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

# Introduction


## Dataset


## Project Objectives

(1). To identify the risk factors for Type 2 diabetes and test the importance of body mass index and age in incident diabetes using data from a cohort of Chinese adults.

(2). To use the joint model to assess the association between survival outcome (death status and event time) with the drug efficacy and the latent mean of CD4 count at event time.

(3). To assess and compare the model fitting result (mean and standard error) for the frequentist and Bayesian methods, as well a machine learning methods. We also simulated similar data to check and compare the model fitting results.

(4). To propose a dynamic prediction framework to patients with longitudinal outcomes available, such as predicting the future longitudinal outcome and conditional survival probability at a given time.

# Methods 



### Model specification



### Likelihood


# Details in methods

We used Ridge-penalized logistic regression, Bayesian methods, and Random Survival Forest to estimate parameters.

### Penalized Ridge logistic regression


### Bayesian methods


### Random Survival Forest


# Software implementation

We write our package `jm5`. Here are the main functions:

```{r eval=FALSE}

```

# Results 


### Summary statistics for `diabetes` dataset

```{r echo=FALSE}

```



### Model fit for `diabetes` dataset - Penalized Ridge regression 

```{r, results = "hide"}

```

It is time-consuming to run MCEM and each iteration takes more than 1 minute. Although the algorithm is hard to converge and we only run 10 iterations, the estimates are still similar with the estimates given by Bayesian method below.

### Model fit for `diabetes` dataset - Bayesian method

```{r warning=FALSE, results="hide"}

```

### Random Forest for `diabetes` dataset

```{r}

```


### Cross-validation: AUC and F1 score for `diabetes` dataset


```{r}

```

# Discussion



## Main conclusion


## Limitations

Below are some advantages and disadvantages of the three methods.

### Penalized Ridge regression 



### Bayesian method


### RSF


## Future direction


# References 

